---
title: "Reproducible Research: Peer Assessment 1"
author: "Martín Dâvila, PhD."
date: "Sunday, October 18, 2015"
output: html_document
---


This assignment makes use of data from a personal activity monitoring device.
This device collects data at 5 minute intervals through out the day.
The data consists of two months of data from an anonymous individual collected during
the months of October and November, 2012 and include the number of steps taken in 5 minute
intervals each day.

# Assignment
This assignment will be described in multiple parts.
You will need to write a report that answers the questions detailed below.
Ultimately, you will need to complete the entire assignment in a single R markdown document
that can be processed by knitr and be transformed into an HTML file.

### Loading and preprocessing the data
1. Load the data (i.e. read.csv())
```{r}
# First of all we need to set the work directory, in which we find all the data sets
# for me is the following
setwd("C:/Coursera/Course Project 1 RR")
getwd()

# Clean up the workspace
rm(list=ls())

# Load the raw activity data
activity_raw <- read.csv("activity.csv", stringsAsFactors=FALSE)
```

2. Process/transform the data (if necessary) into a format suitable for your analysis
```{r}
# Transform the date attribute to an actual date format
activity_raw$date <- as.POSIXct(activity_raw$date, format="%Y-%m-%d")

# Compute the weekdays from the date attribute
activity_raw <- data.frame(date=activity_raw$date, 
                           weekday=tolower(weekdays(activity_raw$date)), 
                           steps=activity_raw$steps, 
                           interval=activity_raw$interval)

# Compute the day type (weekend or weekday)
activity_raw <- cbind(activity_raw, 
                      daytype=ifelse(activity_raw$weekday == "saturday" | 
                                     activity_raw$weekday == "sunday", "weekend", 
                                     "weekday"))

# Create the final data.frame
activity <- data.frame(date=activity_raw$date, 
                       weekday=activity_raw$weekday, 
                       daytype=activity_raw$daytype, 
                       interval=activity_raw$interval,
                       steps=activity_raw$steps)

# Clear the workspace
rm(activity_raw)

# We display the first few rows of the activity data frame:
head(activity)
```

### What is mean total number of steps taken per day?
1. Calculate the total number of steps taken per day
```{r}
# Calculate the total number of steps each day (NA values removed)
sum_data <- aggregate(activity$steps, by=list(activity$date), FUN=sum, na.rm=TRUE)

# Rename the attributes
names(sum_data) <- c("date", "total")

# We display the first few rows of the sum_data data frame:
head(sum_data)
```

2. If you do not understand the difference between a histogram and a barplot,
research the difference between them.
Make a histogram of the total number of steps taken each day.
```{r}
# Draw the histogram of the total number of steps each day
hist(sum_data$total, 
     breaks=seq(from=0, to=25000, by=2500),
     col="blue", 
     xlab="Total number of steps", 
     ylim=c(0, 20), 
     main="Histogram of the total number of steps taken each day\n(NA removed)")
```

3. Calculate and report the mean and median of the total number of steps taken per day.
```{r}
mean(sum_data$total)
median(sum_data$total)
```

### What is the average daily activity pattern?
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis)
and the average number of steps taken, averaged across all days (y-axis)
```{r}
# First of all, we need to clear the workspace again
rm(sum_data)

# Calculate the means of steps accross all days for each interval
mean_data <- aggregate(activity$steps, 
                       by=list(activity$interval), 
                       FUN=mean, 
                       na.rm=TRUE)

# Rename the attributes
names(mean_data) <- c("interval", "mean")

# We display the first few rows of the mean_data data frame:
head(mean_data)

# Draw the time series plot
plot(mean_data$interval, 
     mean_data$mean, 
     type="l", 
     col="blue", 
     lwd=2, 
     xlab="Interval [minutes]", 
     ylab="Average number of steps", 
     main="Time-series of the average number of steps per intervals\n(NA removed)")
```

2. Which 5-minute interval, on average across all the days in the dataset,
contains the maximum number of steps?
```{r}
# We need to find the position of the maximum mean
max_pos <- which(mean_data$mean == max(mean_data$mean))

# We must to lookup the value of interval at this position
max_interval <- mean_data[max_pos, 1]

# Clear again the workspace
rm(max_pos, mean_data)

# Display the answer to the question
print(max_interval)
```

### Imputing missing values
1. Calculate and report the total number of missing values in the dataset
(i.e. the total number of rows with NAs)
```{r}
# Clearing the workspace again
rm(max_interval)

# We need to use the trick that a TRUE boolean value is equivalent to 1 and a FALSE to 0.
NA_count <- sum(is.na(activity$steps))
```

2. Devise a strategy for filling in all of the missing values in the dataset.
The strategy does not need to be sophisticated.
For example, you could use the mean/median for that day,
or the mean for that 5-minute interval, etc.
```{r}
# Again we need to clear the workspace
rm(NA_count)

# We need to find the NA positions
na_pos <- which(is.na(activity$steps))

# We need to create a vector of means
mean_vec <- rep(mean(activity$steps, na.rm=TRUE), times=length(na_pos))
```

3. Create a new dataset that is equal to the original dataset
but with the missing data filled in.
```{r}
# Replace the NAs by the means
activity[na_pos, "steps"] <- mean_vec

# Again we clear the workspace
rm(mean_vec, na_pos)

# We display the first few rows of the new activity data frame:
head(activity)
```

4. Make a histogram of the total number of steps taken each day and
Calculate and report the mean and median total number of steps taken per day.
Do these values differ from the estimates from the first part of the assignment?
What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
# Calculate the total number of steps each day (NA values removed)
sum_data <- aggregate(activity$steps, by=list(activity$date), FUN=sum)

# Rename the attributes
names(sum_data) <- c("date", "total")

# Draw the histogram of the total number of steps each day
hist(sum_data$total, 
     breaks=seq(from=0, to=25000, by=2500),
     col="blue", 
     xlab="Total number of steps", 
     ylim=c(0, 30), 
     main="Histogram of the total number of steps taken each day\n(NA replaced by mean value)")

# The mean and median are computed like
mean(sum_data$total)
median(sum_data$total)
```

### Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend”
indicating whether a given date is a weekday or weekend day.
```{r}
# The new factor variable "daytype" was already in the activity data frame
head(activity)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval
(x-axis) and the average number of steps taken, averaged across all weekday days or weekend days
(y-axis). See the README file in the GitHub repository to see an example of what this plot
should look like using simulated data.
```{r}
# Again we clear the workspace
rm(sum_data)

# We need to load the lattice graphical library
library(lattice)

# Calculate the average number of steps taken, averaged across all daytype variable
mean_data <- aggregate(activity$steps, 
                       by = list(activity$daytype, 
                               activity$weekday, activity$interval), mean)

# Rename the attributes
names(mean_data) <- c("daytype", "weekday", "interval", "mean")

# We display the first few rows of the mean_data data frame:
head(mean_data)

# Draw the time serie plot
xyplot(mean ~ interval | daytype, mean_data, 
       type = "l", 
       lwd = 1, 
       xlab = "Interval", 
       ylab = "Number of steps", 
       layout = c(1,2))

# Again clearing the workspace
rm(mean_data)
```
